---
title: "Homework3"
author: "Devin Teran, Dennis Pong, Richard Zheng, Katie Evers"
date: "10/4/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rmdformats)
library(dplyr)
library(ggplot2)
library(dplyr)
library(kableExtra)
library(gridExtra)
require(ISLR)
require(tree)
library(GGally)
library(mice) # Multivariate Imputation By Chained Equations
library(psych)
library(visdat)
library(tidyr)
library(knitr)
library(rmdformats)
library(tidyverse)
library(caret) # Classification And REgression Training - for createDataPartition, featureplot, classification report, and other key functions to streamline the model training process.
library(corrplot) # For correlation matrix
library(randomForest)
library(MASS)
library(doParallel)
library(parallel)
library(plyr)
library(VIM)

options(max.print="100")
opts_knit$set(width=31)
```

```{r import-data}
# na.strings = "" will treat empty string as NA
loan_data <- read.csv('https://raw.githubusercontent.com/metis-macys-66898/data622_fa2021/main/hw3/data/Loan_approval.csv', header = TRUE, na.strings = "")

loan_raw <- read.csv('https://raw.githubusercontent.com/metis-macys-66898/data622_fa2021/main/hw3/data/Loan_approval.csv', header = TRUE)
```

## 1. Exploratory Data Analysis

Our loan data has 614 rows and 13 columns, 8 of which are categorical and 5 are numerical. The target variable is **Loan\_Status**, which can be either **Y (yes)** or **N (no)**. This let's us know if the applicant's loan is approved. There are 7 variables that have blank values. The one with the most blank values is **Credit\_History** with 50 blanks.

```{r view}
loan_data %>% kbl() %>% kable_styling() %>% scroll_box(width = "750px", height = "250px")

summary(loan_data) %>% kbl() %>% kable_styling() %>% scroll_box(width = "750px", height = "250px")

missing <- loan_data %>% mutate_if(is.character, list(~na_if(.,""))) 

missing%>%
  summarise_all(list(~sum(is.na(.)))) %>%
  gather(key="Variable", value="Number_Missing") %>%
  arrange(desc(Number_Missing)) %>% kbl() %>% kable_styling() %>% scroll_box(width = "750px", height = "250px")
```

### Factors

```{r to-factor}
loan_data <- loan_data %>% mutate_if(is.character, factor)
```

### The following section we'll continue to look at the data from the raw perspective (loan\_raw).

### **Categorical Variables**

There are several variables which have blank values "". These data points may have been intentionally skipped by customers from banks during the data collection process or they may just be missing. We will handle this later on.\
\* **Loan\_ID:** unique identifier\
\* **Gender:** either Male or Female or blank\
\* **Married:** either No or Yes or blank\
\* **Dependents:** how many dependents does someone have? 0, 1, 2, 3+ or blank\
\* **Education:** Graduate or Not Graduate\
\* **Self\_Employed:** No or Yes or blank\
\* **Property\_Area:** Urban, Rural or Semiurban\
\* **Loan\_status:** Y (yes) or N (no)\
\* **Credit\_History:** does the credit history meet the guidelines? 1 = Yes, 0 = No

```{r gender-tables,include=FALSE}
#### Gender  
#More men submitted loan applications than women.  Men have a slightly higher loan approval status than women but not by much. #A small portion of the applications do not list gender.  

gender_loan_status_count <- table(loan_raw$Gender,loan_raw$Loan_Status)
gender_loan_status_perct <- gender_loan_status_count
gender_loan_status_perct[1,] <- round(gender_loan_status_perct[1,]/13 * 100, 2)
gender_loan_status_perct[2,] <- round(gender_loan_status_perct[2,]/112 * 100, 2)
gender_loan_status_perct[3,] <- round(gender_loan_status_perct[3,]/489 * 100, 2)

#set column names for gender_loan_status_count
gender_loan_status_count <- data.frame(gender_loan_status_count)
colnames(gender_loan_status_count) <- c('Gender','Loan_Status','Count')

#set column names and row names for gender_loan_status_perct
rownames(gender_loan_status_perct) <-  c("Blank","Female", "Male")
colnames(gender_loan_status_perct) <-  c("% Applications Not Approved", "% Applications Approved")

loan_data_Gender <- loan_raw
loan_data_Gender[loan_data_Gender$Gender == '',] <- "Blank"

t1 <- loan_data_Gender %>% group_by(Gender) %>% tally 
colnames(t1) <- c("Gender","Count Loan Applications")
t2 <- gender_loan_status_perct

knitr::kable(list(t1, t2))
```

```{r gender-histogram,include = FALSE}
ggplot(data=gender_loan_status_count, aes(x=Gender, y=Count, fill=Loan_Status)) + geom_bar(stat="identity",position="dodge")
```

#### Married

Married applicants have a higher approval rate than non married applicants. It will be useful to look into if this has any correlation with income.

```{r married-tables,fig.width=7,fig.height=3}
married_loan_status_count <- table(loan_raw$Married,loan_raw$Loan_Status)
married_loan_status_perct <- married_loan_status_count
married_loan_status_perct[1,] <- round(married_loan_status_perct[1,]/3 * 100, 2)
married_loan_status_perct[2,] <- round(married_loan_status_perct[2,]/213 * 100, 2)
married_loan_status_perct[3,] <- round(married_loan_status_perct[3,]/398 * 100, 2)

#set column names for married_loan_status_count
married_loan_status_count <- data.frame(married_loan_status_count)
colnames(married_loan_status_count) <- c('Married','Loan_Status','Count')

#set column names and row names for gender_loan_status_perct
rownames(married_loan_status_perct) <-  c("Blank", "Not Married", "Married")
colnames(married_loan_status_perct) <-  c("% Applications Not Approved", "% Applications Approved")

loan_data_Married <- loan_raw
loan_data_Married[loan_data_Married$Married == '',] <- "Blank"

t1 <- loan_data_Married %>% group_by(Married) %>% tally 
colnames(t1) <- c("Married","Count Loan Applications")
t2 <- married_loan_status_perct

knitr::kable(list(t1, t2)) 

```

```{r married-histogram,fig.width=7,fig.height=3}
ggplot(data=married_loan_status_count, aes(x=Married, y=Count, fill=Loan_Status)) + geom_bar(stat="identity",position="dodge")
```

#### Dependents

Applicants with 2 dependents appear to have the highest loan approval rate. It'd be interesting to see if the income per dependent has any impact on loan approval if we assume having more income makes it more likely to get a loan approved.

```{r dep-tables,fig.width=7,fig.height=3}
dep_loan_status_count <- table(loan_raw$Dependents,loan_raw$Loan_Status)
dep_loan_status_perct <- dep_loan_status_count
dep_loan_status_perct[1,] <- round(dep_loan_status_perct[1,]/15 * 100, 2)
dep_loan_status_perct[2,] <- round(dep_loan_status_perct[2,]/345 * 100, 2)
dep_loan_status_perct[3,] <- round(dep_loan_status_perct[3,]/102 * 100, 2)
dep_loan_status_perct[4,] <- round(dep_loan_status_perct[4,]/101 * 100, 2)
dep_loan_status_perct[5,] <- round(dep_loan_status_perct[5,]/51 * 100, 2)

#set column names for dep_loan_status_count
dep_loan_status_count <- data.frame(dep_loan_status_count)
colnames(dep_loan_status_count) <- c('Dependents','Loan_Status','Count')

#set column names and row names for gender_loan_status_perct
rownames(dep_loan_status_perct) <-  c("Blank", "0", "1","2","3+")
colnames(dep_loan_status_perct) <-  c("% Applications Not Approved", "% Applications Approved")

loan_data_Dep <- loan_raw
loan_data_Dep[loan_data_Dep$Dependents == '',] <- "Blank"

t1 <- loan_data_Dep %>% group_by(Dependents) %>% tally 
colnames(t1) <- c("Dependents","Count Loan Applications")
t2 <- dep_loan_status_perct

knitr::kable(list(t1, t2))
```

```{r dep-histogram,fig.width=7,fig.height=3}
ggplot(data=dep_loan_status_count, aes(x=Dependents, y=Count, fill=Loan_Status)) + geom_bar(stat="identity",position="dodge")
```

#### Education

Applicants with Graduate education have a higher loan approval rate here.

```{r edu-tables,fig.width=7,fig.height=3}
edu_loan_status_count <- table(loan_raw$Education,loan_raw$Loan_Status)
edu_loan_status_perct <- edu_loan_status_count
edu_loan_status_perct[1,] <- round(edu_loan_status_perct[1,]/480 * 100, 2)
edu_loan_status_perct[2,] <- round(edu_loan_status_perct[2,]/134 * 100, 2)

#set column names for edu_loan_status_count
edu_loan_status_count <- data.frame(edu_loan_status_count)
colnames(edu_loan_status_count) <- c('Education','Loan_Status','Count')

#set column names for edu_loan_status_perct
colnames(edu_loan_status_perct) <-  c("% Applications Not Approved", "% Applications Approved")

t1 <- loan_raw %>% group_by(Education) %>% tally 
colnames(t1) <- c("Education","Count Loan Applications")
t2 <- edu_loan_status_perct

knitr::kable(list(t1, t2))
```

```{r edu-histogram,fig.width=7,fig.height=3}
ggplot(data=edu_loan_status_count, aes(x=Education, y=Count, fill=Loan_Status)) + geom_bar(stat="identity",position="dodge")
```

```{r selfemp-tables,include=FALSE}
#### Self Employed    
#The loan approval rate is almost identical here for self employed versus not self employed.  

selfemp_loan_status_count <- table(loan_raw$Self_Employed,loan_raw$Loan_Status)
selfemp_loan_status_perct <- selfemp_loan_status_count
selfemp_loan_status_perct[1,] <- round(selfemp_loan_status_perct[1,]/32  * 100, 2)
selfemp_loan_status_perct[2,] <- round(selfemp_loan_status_perct[2,]/500  * 100, 2)
selfemp_loan_status_perct[3,] <- round(selfemp_loan_status_perct[3,]/82  * 100, 2)

#set column names for selfemp_loan_status_count
selfemp_loan_status_count <- data.frame(selfemp_loan_status_count)
colnames(selfemp_loan_status_count) <- c('Self_Employed','Loan_Status','Count')

#set column names and row names for selfemp_loan_status_perct
rownames(selfemp_loan_status_perct) <-  c("Blank", "No", "Yes")
colnames(selfemp_loan_status_perct) <-  c("% Applications Not Approved", "% Applications Approved")

loan_data_selfemp<- loan_raw
loan_data_selfemp[loan_data_selfemp$Self_Employed == '',] <- "Blank"
t1 <- loan_data_selfemp %>% group_by(Self_Employed) %>% tally 
colnames(t1) <- c("Self_Employed","Count Loan Applications")
t2 <- selfemp_loan_status_perct

knitr::kable(list(t1, t2))
```

```{r self-employed-histogram,include=FALSE}
ggplot(data=selfemp_loan_status_count, aes(x=Self_Employed, y=Count, fill=Loan_Status)) + geom_bar(stat="identity",position="dodge")
```

#### Property Area

Semiurban applicants have the highest approval loan rating over rural and urban.

```{r proparea-tables,fig.width=7,fig.height=3}
proparea_loan_status_count <- table(loan_raw$Property_Area,loan_raw$Loan_Status)
proparea_loan_status_perct <- proparea_loan_status_count
proparea_loan_status_perct[1,] <- round(proparea_loan_status_perct[1,]/179  * 100, 2)
proparea_loan_status_perct[2,] <- round(proparea_loan_status_perct[2,]/233  * 100, 2)
proparea_loan_status_perct[3,] <- round(proparea_loan_status_perct[3,]/202  * 100, 2)

#set column names for proparea_loan_status_count
proparea_loan_status_count <- data.frame(proparea_loan_status_count)
colnames(proparea_loan_status_count) <- c('Property_Area','Loan_Status','Count')

#set column names for proparea_loan_status_perct
colnames(proparea_loan_status_perct) <-  c("% Applications Not Approved", "% Applications Approved")

t1 <- loan_raw %>% group_by(Property_Area) %>% tally 
colnames(t1) <- c("Property_Area","Count Loan Applications")
t2 <- proparea_loan_status_perct

knitr::kable(list(t1, t2))
```

```{r property-area-histogram,fig.width=7,fig.height=3}
ggplot(data=proparea_loan_status_count, aes(x=Property_Area, y=Count, fill=Loan_Status)) + geom_bar(stat="identity",position="dodge")
```

#### Credit History

Having an a credit history that meets the guidelines appears to be extremely important in whether the loan status is approved or not.

```{r credit-history-tables,fig.width=7,fig.height=3}
credhist_loan_status_count <- table(loan_raw$Credit_History,loan_raw$Loan_Status)
credhist_loan_status_perct <- credhist_loan_status_count
credhist_loan_status_perct[1,] <- round(credhist_loan_status_perct[1,]/89  * 100, 2)
credhist_loan_status_perct[2,] <- round(credhist_loan_status_perct[2,]/475  * 100, 2)

#set column names for credhist_loan_status_count
credhist_loan_status_count <- data.frame(credhist_loan_status_count)
colnames(credhist_loan_status_count) <- c('Credit_History','Loan_Status','Count')

#set column names for credhist_loan_status_perct
colnames(credhist_loan_status_perct) <-  c("% Applications Not Approved", "% Applications Approved")

t1 <- loan_raw %>% group_by(Credit_History) %>% tally 
colnames(t1) <- c("Credit_History","Count Loan Applications")
t2 <- credhist_loan_status_perct

knitr::kable(list(t1, t2))
```

```{r credit_history-histogram,fig.width=7,fig.height=3}
ggplot(data=credhist_loan_status_count, aes(x=Credit_History, y=Count, fill=Loan_Status)) + geom_bar(stat="identity",position="dodge")
```

### **Numerical Variables**

-   **ApplicantIncome:** how much money does the applicant make?\
-   **CoapplicantIncome:** how much money does the coapplicant make? if there is no coapplicant this is 0.\
-   **LoanAmount:** how much is the loan worth in thousands?\
-   **Loan\_Amount\_Term:** how many months is the loan?

Now let's use the **pairs.panels** function to see a lot of important information related to our numeric data:

-   **Applicant income** and **loan\_amount** are strongly correlated\
-   The most common **Loan\_Amount\_Term** is 360 months

```{r summary}
numeric_loan_data <- dplyr::select(loan_data,ApplicantIncome,CoapplicantIncome,LoanAmount,Loan_Amount_Term)
pairs.panels(numeric_loan_data, 
             method = "pearson", # correlation method
             hist.col = "#00AFBB",
             density = TRUE,  # show density plots
             ellipses = TRUE # show correlation ellipses
             )
```

#### Inspecting ApplicantIncome and Loan Income

Here we can see that the **ApplicantIncome** does not have a huge effect on whether the Loan\_Status was approved (Y) or not. The average ApplicantIncome is about the same for both groups is similar. There are a fewer more outliers of high incomes in the group where the loan status was approved.

```{r plot-income, warning = FALSE,fig.width=7,fig.height=6}
approved <- loan_data[loan_data$Loan_Status == 'Y',]
denied <- loan_data[loan_data$Loan_Status == 'N',]

a <- ggplot(loan_data,aes(x=ApplicantIncome,color=Loan_Status))  + geom_boxplot()
b <- ggplot(approved,aes(x=ApplicantIncome,y=LoanAmount,color=Loan_Status)) + geom_point(color='blue') + xlab('Approved Applicant Income') + scale_x_continuous(limits = c(0, 25000)) + scale_y_continuous(limits = c(0, 650))
grid.arrange(a,b,nrow=2)#,nrow=2,ncol=2,layout_matrix=c(1,1,2,3)) 
```

```{r plot-income-1,warning = FALSE,fig.width=7,fig.height=3}
c <- ggplot(denied,aes(x=ApplicantIncome,y=LoanAmount,color=Loan_Status)) + geom_point(color='red') + xlab('Denied Applicant Income') + scale_x_continuous(limits = c(0, 25000)) + scale_y_continuous(limits = c(0, 650))

grid.arrange(c)

```

In addition, upon investigating the sum of **ApplicantIncome** and **CoapplicantIncome**, we observe that it does not appear to have much prediction power with **Loan\_Status**.

```{r plot-income-combined,warning = FALSE,fig.width=7,fig.height=3}
ggplot(data = loan_data, aes(x = Loan_Status, y = ApplicantIncome+CoapplicantIncome, fill=Loan_Status)) +
  geom_boxplot() +
  coord_flip()
```

#### LoanAmount Per ApplicantIncome

Now let's see if the rate of the **LoanAmount** divided by **ApplicantIncome** has any prediction power when trying to deteremine if a Loan\_Status will be approved or not. This would indicate that perhaps someone who is requesting a LoanAmount 5 times their income, they might not be approved but if they requested 3 times their income they could get approved.

Looking at the boxplots below, the average LoanAmtPerSalary is roughly the same for approved and not approved applications so this disbunks this theory. This variable might prove helpful in our modeling so we will keep it.

```{r plot-loan-per-income,fig.width=7,fig.height=3,warnings=FALSE}
loan_data$LoanAmtPerSalary <- loan_data$LoanAmount*100000/loan_data$ApplicantIncome
ggplot(loan_data,aes(x=LoanAmtPerSalary,color=Loan_Status)) + geom_boxplot() + scale_x_continuous(limits = c(0, 30000))


```

## Data Prep for Model-fitting

I explicitly recoded the Y/N values into 1/0's

```{r data-prep, include=FALSE}
loan_knn <- loan_data
loan_knn$Loan_Status <- as.numeric(loan_knn$Loan_Status)-1 
str(loan_knn)
```

Since credit history is a categorical value and fewer than 50 rows are missing it's better to delete these data points rather than to try to interpret a value for them. For loan amount term and loan amount we will use the **mice** package to impute a value where it is missing.

## Additional Data Processing / Manipulation Steps

So, first off, I need to convert Credit\_History to factors so that the mice model that I'm going to use can detect that column as a categorical variable.

Combining ApplicantIncome and CoapplicantIncome into a new variable TotalIncome, and dropping the respective input columns. Loan\_ID doesn't help with the prediction obviously. So dropping it as well.

```{r addl-data-process}

loan_knn_pre_imp <- loan_knn
loan_knn_pre_imp$Credit_History <- as.factor(loan_knn_pre_imp$Credit_History)

loan_knn_pre_imp <- loan_knn_pre_imp %>% mutate(TotalIncome = ApplicantIncome + CoapplicantIncome)
loan_knn_pre_imp <- loan_knn_pre_imp %>% dplyr::select(-c('Loan_ID','ApplicantIncome','CoapplicantIncome'))

# loan_knn_pre_imp[loan_knn_pre_imp$Dependents = "3+"] <- "3"

# recode dependents 3+ to 3
loan_knn_pre_imp$Dependents <- revalue(loan_knn_pre_imp$Dependents, c("3+"="3"))


str(loan_knn_pre_imp)
```

### Status quo of missing data

```{r missing-counts,include=TRUE}

clean_loan_data <- loan_data

vis_dat(clean_loan_data)
```

I've set up a predictorMatrix where I can instruct mice to use which method for which column for imputation.

Set seed = 501. Retrieved the results.

```{r impute-missing-numeric-data,class.source = 'fold-show',warnings=FALSE}

# clean_loan_data <- complete(mice(clean_loan_data,m=5,meth='pmm',print=FALSE))
init <- mice(loan_knn_pre_imp, maxit=0) 
meth <- init$method
predM <- init$predictorMatrix
meth[c('LoanAmount','Loan_Amount_Term')] <- 'norm'
meth[c('Credit_History','Self_Employed','Gender','Married')] <- 'logreg'
meth[c('Dependents')] <- 'polyreg'
meth[c('Loan_Status','TotalIncome','Property_Area','Education')] = ''
loan_knn_imp1 <- mice(loan_knn_pre_imp, method=meth, predictorMatrix=predM, seed=501)

```

### Manual Examinations

After some manual examinations of the different imputed results, I've decided to go with imputed column \#3.

```{r manual-exams}
# Manual examination 
#Credit_History
loan_knn_imp1$imp$Credit_History
loan_knn[96:118,]

#Married
loan_knn_imp1$imp$Married
loan_knn[430:436,]

#Dependents
loan_knn_imp1$imp$Dependents
loan_knn[227:229,]

```

### Decision

We picked impute \#3.

```{r decision}
loan_knn2 <- complete(loan_knn_imp1, 3) # 2nd argument if not provided is defaulted to 1
```

```{r clean_loan_data}
clean_loan_data <- loan_knn2

# have to redo loan_status as loan_knn's loan status had been recoded to numeric on purpose
clean_loan_data$Loan_Status <- as.factor(clean_loan_data$Loan_Status)
str(clean_loan_data)
```

## Imbalanced Dataset

Notice that the response variable is 31/69 split on the binary response, No and Yes, respectively.

```{r imbalanced-data}
imb_dat <- as.data.frame(prop.table(x = table(clean_loan_data$Loan_Status)))
colnames(imb_dat) <- c("Loan Status", "Freq")
imb_dat

```

### Splitting Data into Training & Testing

Here we are going to use 80% of our data to train the model and reserve 20% to test the model we pick.

```{r train-test}
set.seed(1042)
sample_size <- floor(nrow(clean_loan_data)*0.8)
indices <- sample(1:nrow(clean_loan_data),sample_size)
train <- clean_loan_data[c(indices),]
test <- clean_loan_data[-c(indices),]
```

## 2. Linear Discriminant Analysis

LDA does not seem to be a good approach with this data set as the points provided by the available data are not linearly separable

```{r}
train%>%
  ggplot(aes(x = log(LoanAmount), y= log(TotalIncome), color = Loan_Status)) + geom_point()
```

### lda Cross Validation

predictions with the LDA model are less accurate than if we just used the binary classifier Credit\_History to determine weather or not a loan would be approved

```{r trainCtrl}
# cross validation 
ctrl <- trainControl(method = 'repeatedcv', repeats = 11)
```

### lda model results

```{r}
lda.fit <- train(Loan_Status ~ TotalIncome + LoanAmount,
             data = train,
             method = 'lda',
             trControl = ctrl
             )
test$lda <- predict(lda.fit, test)
confusionMatrix(test$lda, test$Loan_Status)
```

## 3. K-nearest Neighbor

First off, set seed = 688.

Create training/test partitions by calling createDataPartition. p is set to .8 to mean 80/20 split for train/test set.

Checking the structure of the train set (*knn\_train*)

```{r knn_train, include=FALSE}
set.seed(688)
# recoding Loan_Status back to categorical variable
loan_knn2$Loan_Status <- as.factor(loan_knn2$Loan_Status)
str(loan_knn2)


# Data Partitioning
trainIndex <- createDataPartition(loan_knn2$Loan_Status, p = .8, list = FALSE, times = 1)
knn_train <- loan_knn2[trainIndex,]
knn_test  <- loan_knn2[-trainIndex,]



str(knn_train)
```

Checking the structure of the test set (*knn\_test*)

```{r knn_test}
str(knn_test)
```

### Cross Validation

Perform a repeated 11-fold cross-validation, meaning the number of complete sets of folks to compute is 11. For this classification problem, we assigned our fitted model to *knn.fit.* The cross-validated results is plugged in the form of *trControl.*

```{r knn_fit}

# cleaning up some parallel computing
# https://stackoverflow.com/questions/25097729/un-register-a-doparallel-cluster
registerDoSEQ()

trControl <- trainControl(method  = "repeatedcv",
                          repeats  = 11)
knn.fit <- train(Loan_Status ~ .,
             method     = "knn",
             tuneGrid   = expand.grid(k = 1:10),
             trControl  = trControl,
             preProcess = c("center","scale"),
             data       = knn_train
             )

knn.fit 
```

#### Key Information on the value of K

Since our target variable is a binary factor of 2, by default, we use Accuracy as the determining performance metric. The optimal K is thus determined by Accuracy. **K = 9 was finally selected.** \# of neighbors is 9.

```{r plot knn.fit}
plot(knn.fit)
```

### Model Results

```{r knn-confusion-mat}

knn_pred <- predict(knn.fit, newdata = knn_test)
# options('max.print' = 100)  
# getOption("max.print")
confusionMatrix(knn_pred, knn_test$Loan_Status)

```

Accuracy is 78.69% while balanced accuracy is only 67.95%.

## 4. Decision Trees

Now we will use a decision tree to see how well it will perform on our data.\
\* Our decision tree starts by splitting users based on their Credit\_History. This makes sense based on our exploratory data analysis.\
\* Other variables used in the decision tree include LoanAmount, PropertyArea, etc.

```{r decision-tree,fig.width=6,fig.height=6}
loan_tree = tree(Loan_Status ~., train)
plot(loan_tree)
text(loan_tree)
title(main = "Unpruned Decision Tree")
```

### Decision Tree Performance

#### Training Data

Now we will use our model to see how it performs on the training data. We see that the model predicted **Loan\_Status** with an accuracy of \~83%. 81 instances were incorrectly classified.

```{r tree-test}
pred_tree_train <- predict(loan_tree,train,type="class")
test_table <- table(pred_tree_train,train$Loan_Status) %>% kbl() %>% kable_styling()
test_table
mean(pred_tree_train == train$Loan_Status)
```

### Cross-validation for better performance

The first version of our model was a full, unpruned tree. Now we are going to prune it back to get the optimal tree using cross validation. We have plotted the number of misclassifications with the different trees. As we can see, the trees with size 2-4 have the fewest misclassifications. We will choose size 4 to have the fewest misclassifications.

```{r cv-trees}
set.seed(2311)
cv_trees = cv.tree(loan_tree,FUN = prune.misclass)
cv_trees
plot(cv_trees)
```

Using a size = 4, our decision tree looks like the following:

```{r pruned-tree,figure.height=8}
loan_tree_pruned = prune.misclass(loan_tree,best=4)
plot(loan_tree_pruned)
text(loan_tree_pruned)
```

### Testing Data

Now let's see how our pruned performs on our testing data. The accuracy for our test data was \~82%, which was almost the same as our training data. 21 of the total observations were misclassified.

```{r confusion-matrix}
pred_tree_test  <- predict(loan_tree_pruned,test, type="class")
test_table <- table(pred_tree_test,test$Loan_Status) %>% kbl() %>% kable_styling()
test_table
mean(pred_tree_test == test$Loan_Status)

confusionMatrix(pred_tree_test, test$Loan_Status)
```

## 5. Random Forests

Now we will develop a random forest model to see how well it will performs with our data. Parameters for a random forest include: **mtry** : Number of variables randomly sampled as candidates at each split. Note that the default values are different for classification (sqrt(p) where p is number of variables in x) and regression (p/3) **ntree** : Number of trees to grow. This should not be set to too small a number, to ensure that every input row gets predicted at least a few times

Our initial model will have default parameters of mtry=sqrt(13) and ntree=500.

```{r rf-start-step1}
# find out no of cores 
no_cores <- detectCores() - 1

cl<-makePSOCKcluster(no_cores)
  
registerDoParallel(cl)
  
# start.time<-proc.time()
  
# model<-train(target~., data=trainingset, method='rf')
```

```{r rf-default-step2}
#drop loan id
train_rf1 <- train
test_rf1 <- test 

# Create model with default parameters
control <- trainControl(method="repeatedcv", number=10, repeats=3)
mtry <- sqrt(ncol(train_rf1))
tunegrid <- expand.grid(.mtry=mtry)
rf_default <- train(Loan_Status~., data=train_rf1, method="rf", metric="Accuracy", tuneGrid=tunegrid, trControl=control)
print(rf_default)

# stop.time<-proc.time()
# 
# run.time<-stop.time -start.time
# 
# print(run.time)
#   
# stopCluster(cl)
```

Our inital model has accuracy of about 80%. Let's see if we can improve accuracy by finding an optimal mtry value. We will test different mtry values 1-10 by using gridsearch. We see from our results that the optimal mtry value for accuracy is 2.

```{r rf-mtry}
control <- trainControl(method="repeatedcv", number=10, repeats=3, search="grid")
set.seed(123)
tunegrid <- expand.grid(.mtry=c(1:10))
rf_gridsearch <- train(Loan_Status~., data=train_rf1, method="rf", metric="Accuracy", tuneGrid=tunegrid, trControl=control)
print(rf_gridsearch)
plot(rf_gridsearch)
```

Next let's find the optimal value for ntree. Again we'll use gridsearch to test different ntree values. It's evident from our results that optimal ntree value for accuracy is 1500.

```{r rf-ntree}
control <- trainControl(method="repeatedcv", number=10, repeats=3, search="grid")
tunegrid <- expand.grid(.mtry=2)
modellist <- list()
for (ntree in c(500, 1000, 1500, 2000, 2500)) {
  set.seed(124)
	fit <- train(Loan_Status~., data=train_rf1, method="rf", metric="Accuracy", tuneGrid=tunegrid, trControl=control, ntree=ntree)
	key <- toString(ntree)
	modellist[[key]] <- fit
}
# compare results
results <- resamples(modellist)
summary(results)
dotplot(results)
```

Our final random forest model will have mtry=2 and ntree=1500.

```{r rf-final}
rf_final <- randomForest(Loan_Status ~ ., 
                        data = train_rf1, 
                        ntree = 1500, 
                        mtry = 2,
                        importance = TRUE,
                        proximity = TRUE)

print(rf_final)
#variable importance
round(importance(rf_final), 2)

prediction <-predict(rf_final, test_rf1)
confusionMatrix(prediction, test_rf1$Loan_Status)

# stop.time<-proc.time()

# run.time<-stop.time -start.time

# print(run.time)

# Stopping Cluster
stopCluster(cl)
```

Accuracy of our final random forest model is about 83% on the test data with 19 instances misclassified. 2 are false negatives and 17 are false positives. Credit\_history is the most important feature.

## 6. Model Performance

|       Metric      |   LDA  | K-Nearest Neighbor (KNN) | Decision Trees | Random Forest |
|:-----------------:|:------:|:------------------------:|:--------------:|:-------------:|
|      Accuracy     | 0.7317 |          0.7869          |     0.8293     |     0.8374    |
| Balanced Accuracy | 0.5000 |          0.6795          |     0.7010     |     0.7066    |
|    Sensitivity    |    0   |          0.3947          |     0.4242     |     0.4242    |

: Model Performance Matrix

Notice that the sensitivity between Decision Trees, and RF is the same at 42.42%. It's surprising to see that LDA and Random Forest ended up having the highest accuracy, which is usually the go-to metric to go for in an unbalanced dataset with the binary response that are not 50/50. The model we picked is Random Forest.

### Resources:

-   <https://www.r-bloggers.com/2015/10/imputing-missing-data-with-r-mice-package/>\
-   <https://daviddalpiaz.github.io/r4sl/trees.html>
-   <https://www.youtube.com/watch?v=YPz2J5lHeVM>
-   <https://machinelearningmastery.com/tune-machine-learning-algorithms-in-r/>
