---
title: "DATA 622 HW3"
author: "Dennis Pong"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
    code_folding: hide
---

```{r setup, cache=FALSE}
library(knitr)
library(rmdformats)
library(tidyverse)
library(caret) # For createDataPartition, featureplot, classification report
library(corrplot) # For correlation matrix
library(mice) # Multivariate Imputation By Chained Equations
library(parallel)
library(plyr)
library(VIM)

## Global options


options(max.print="100")
opts_knit$set(width=31)
```

# **K-Nearest Neighbors**

```{r loading-data}
loan_raw <- read.csv('https://raw.githubusercontent.com/metis-macys-66898/data622_fa2021/main/hw3/data/Loan_approval.csv', header = TRUE, na.strings = " ")
loan_raw[loan_raw==""] <- NA
loan_raw <- loan_raw %>% mutate_if(is.character, factor)
loan <- loan_raw
```

## Processing

Besides creating a data.frame called loan\_knn, I explicitly recoded the Y/N values into 1/0's.

```{r}
str(loan)

loan_knn <- loan
loan_knn$Loan_Status <- as.numeric(loan_knn$Loan_Status)-1 
str(loan_knn)



```

Besides, I also used colSums to look for how many rows of missing values for each column in the loan\_knn data.frame.

```{r}
colSums(is.na(loan_knn))

```

After that, I plotted a histogram to better illustrate where are the missing values coming from. It's essentially another way of looking at the distribution of missing values.

```{r}
aggr_plot <- aggr(loan_knn, col=c('navyblue','red'), numbers=TRUE, sortVars=TRUE, labels=names(data), cex.axis=.7, gap=3, ylab=c("Histogram of missing data","Pattern"))


```

There is a total of 6 columns that require imputation. They can be triaged into 4 different categories.

Below we showed you where each of the 6 columns fall under and follow by the eventual method we used in the mice package. Mice package implements a method to deal with missing data. It's short for Multivariate Imputation by Chained Equations.

**Categorical Variables with more than 2 levels:** Dependents(4) - ~~polyreg~~ -\> pmm

**Categorical Variables with 2 levels:** Gender (2), Married(3), Credit\_History(11), Self\_Employed (6) - logreg

**Discrete variable (ordered) with more than 2 levels:** Loan\_Amount\_Term (10) - polr

**Continuous Variables:** LoanAmount (9) - pmm

## Additional Data (Processing) Manipulation Steps

So, first off, I need to convert Credit\_History to factors so that the mice model that I'm going to use can detect that column as a categorical variable.

Combining ApplicantIncome and CoapplicantIncome into a new variable TotalIncome, and dropping the respective input columns. Loan\_ID doesn't help with the prediction obviously. So dropping it as well.

```{r}

loan_knn_pre_imp <- loan_knn
loan_knn_pre_imp$Credit_History <- as.factor(loan_knn_pre_imp$Credit_History)



loan_knn_pre_imp <- loan_knn_pre_imp %>% mutate(TotalIncome = ApplicantIncome + CoapplicantIncome)
loan_knn_pre_imp <- loan_knn_pre_imp %>% select(-c('Loan_ID','ApplicantIncome','CoapplicantIncome'))

# loan_knn_pre_imp[loan_knn_pre_imp$Dependents = "3+"] <- "3"

# recode dependents 3+ to 3
loan_knn_pre_imp$Dependents <- revalue(loan_knn_pre_imp$Dependents, c("3+"="3"))


str(loan_knn_pre_imp)
```

I've set up a predictorMatrix where I can instruct mice to use which method for which column for imputation.

Set seed = 501. Retrieved the results.

```{r mice}
# loan_knn_imp1 <- mice (loan_knn_pre_imp, method = c("",
#                                                     "logreg",
#                                                     "logreg",
#                                                     "polyreg",
#                                                     "",
#                                                     "logreg",
#                                                     "",
#                                                     "",
#                                                     "cart",
#                                                     "polr",
#                                                     "logreg",
#                                                     "",
#                                                     ""
#                                                    ), seed=501, nnet.MaxNWts = 30100
# )
# 
# loan_knn_imp1 <- mice (loan_knn_pre_imp, seed=501, nnet.MaxNWts = 30100)
# 
# loan_knn_imp2 <- mice (loan_knn_pre_imp, method = "rf", seed=501, nnet.MaxNWts = 30100)


# loan_knn_imp2 <- parlmice(data = loan_knn_pre_imp, method = c(  "", 
#                                                                 "logreg", 
#                                                                 "logreg", 
#                                                                 "polyreg", 
#                                                                 "", 
#                                                                 "", 
#                                                                 "", 
#                                                                 "", 
#                                                                 "cart",
#                                                                 "polr",
#                                                                 "logreg",
#                                                                 "",
#                                                                 ""
#                                                               ), cluster.seed = 501, nnet.MaxNWts = 30100, n.core = 2, n.imp.core = 150)



init <- mice(loan_knn_pre_imp, maxit=0) 
meth <- init$method
predM <- init$predictorMatrix
# meth[c('Loan_Amount_Term')] <- 'polr'
# meth[c('LoanAmount')] <- 'cart' 
meth[c('LoanAmount','Loan_Amount_Term')] <- 'norm'
meth[c('Credit_History','Self_Employed','Gender','Married')] <- 'logreg'
meth[c('Dependents')] <- 'polyreg'
meth[c('Loan_Status','TotalIncome','Property_Area','Education')] = ''
loan_knn_imp1 <- mice(loan_knn_pre_imp, method=meth, predictorMatrix=predM, seed=501)




```

After some manual examinations of the different imputed results, I've decided to go with imputed column \#3.

```{r}
# Manual examination 
#Credit_History
loan_knn[96:118,]
loan_knn_imp1$imp$Credit_History
#Married
loan_knn[430:436,]
loan_knn_imp1$imp$Married
#Dependents
loan_knn[227:229,]
loan_knn_imp1$imp$Dependents


loan_knn2 <- complete(loan_knn_imp1, 3) # 2nd argument if not provided is defaulted to 1
```

## **Modeling with KNN**

First off, set seed = 688.

Create training/test partitions by calling createDataPartition. p is set to .8 to mean 80/20 split for train/test set.

Checking the structure of the train set (*knn\_train*)

```{r}
set.seed(688)
# recoding Loan_Status back to categorical variable
loan_knn2$Loan_Status <- as.factor(loan_knn2$Loan_Status)
str(loan_knn2)


# Data Partitioning
trainIndex <- createDataPartition(loan_knn2$Loan_Status, p = .8, list = FALSE, times = 1)
knn_train <- loan_knn2[trainIndex,]
knn_test  <- loan_knn2[-trainIndex,]



str(knn_train)
```

Checking the structure of the test set (*knn\_test*)

```{r}
str(knn_test)
```

### Cross Validation

Perform a repeated 11-fold cross-validation, meaning the number of complete sets of folks to compute is 11. For this classification problem, we assigned our fitted model to *knn.fit.* The cross-validated results is plugged in the form of *trControl.*

```{r}
trControl <- trainControl(method  = "repeatedcv",
                          repeats  = 11)
knn.fit <- train(Loan_Status ~ .,
             method     = "knn",
             tuneGrid   = expand.grid(k = 1:10),
             trControl  = trControl,
             preProcess = c("center","scale"),
             data       = knn_train
             )
```

Since our target variable is a binary factor of 2, by default, we use Accuracy as the determining performance metric. The optimal K is thus determined by Accuracy. **K = 9 was finally selected.** \# of neighbors is 9.

```{r}
# getOption("max.print")
knn.fit
```

```{r echo=FALSE}
plot(knn.fit)
```

Accuracy is 77.1% while balanced accuracy is only 67.48%.

```{r}

knn_pred <- predict(knn.fit, newdata = knn_test)
# options('max.print' = 100)  
# getOption("max.print")
confusionMatrix(knn_pred, knn_test$Loan_Status)


```
